The Galactic Delay-Time Distribution of Type Ia Supernovae:\\ A Chemical Evolution Perspective

### 2023-08-30 src/tex/ms.tex
[showyourwork] overleaf sync

\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{layouts}
\usepackage{xcolor}
\usepackage{upgreek}
% \usepackage{siunitx}

% custom unit definitions
% \DeclareSIUnit\year{yr}

@@

@@

\begin{figure}
    \centering
    \includegraphics{figures/delay_time_distributions.pdf}
    \caption{Models of the SN Ia delay time distribution used in this paper. All functions are normalized to $f_{\rm Ia}=1$ at $\tau=1$ Gyr. The black squares represent the DTD recovered for the SDSS-II sample of SNe Ia by \citet{Maoz2012-SloanIIDTD}. The horizontal error bars indicate the time range of each DTD measurement, and the vertical error bars represent the 1$\sigma$ uncertainties.}
    \label{fig:dtds}
    \script{delay_time_distributions.py}
\end{figure}

@@

\paragraph{Plateau} A ``plateau'' model consisting of a flat slope of width $W$ followed by a power law decline with normalization $B_W$:
\begin{equation}
    R(t) = 
    \begin{cases}
        B_W, & t < W \\
        B_W (t/W)^\alpha, & t \ge W
    \end{cases}
    \label{eq:plateau-dtd}
\end{equation}
where $B_W=[(W-t_{\rm min}) + \frac{t_{\rm max}^{\alpha+1} - W^{\alpha+1}}{(\alpha+1)W^\alpha}]^{-1}$ is the normalization coefficient.
We investigate the cases $W=0.3$ Gyr and $W=1$ Gyr, taking $\alpha=-1.1$ for all plateau models. Figure \ref{fig:dtd-greggio05} illustrates the similarity between these models and two different treatments of the analytical double-degenerate DTD from \citet{Greggio2005-AnalyticalRates}.

@@

\paragraph{Two-population} A DTD in which 50\% of SNe Ia belong to a ``prompt'' Gaussian component at small $t$ and the other 50\% form an exponential tail at large $t$, normalized by $D_{t_p,\sigma,\tau}$:
\begin{equation}
    R(t) = D_{t_p,\sigma,\tau} \Big[e^{-\frac{(t-t_p)^2}{2\sigma^2}} + e^{-t/\tau}\Big]
    \label{eq:prompt-dtd}
\end{equation}
Following the two-population model from \citet{Mannucci2006-TwoPopulations}, we take $t_p=50$ Myr, $\sigma=15$ Myr, and $\tau=3$ Gyr, which results in roughly 50\% of SNe Ia exploding within $t<100$ Myr.

@@

\begin{figure*}
    \centering
    \includegraphics{figures/star_formation_histories.pdf}
    \caption{The surface densities of star formation $\dot \Sigma_*$ (first row), gas infall $\dot \Sigma_{\rm in}$ (second row), and gas mass $\Sigma_{\rm gas}$ (third row), and the star formation efficiency timescale $\tau_*$ (fourth row) as functions of simulation time for our four model SFHs: inside-out (first column; see Equation \ref{eq:insideout-sfh}), late-burst (second column; see Equation \ref{eq:lateburst-sfh}), early-burst (third column; see Equations \ref{eq:earlyburst-ifr} and \ref{eq:earlyburst-taustar}), and two-infall (fourth column; see Equation \ref{eq:twoinfall-ifr}). In each panel, we plot curves for the zones which have inner radii at 4 kpc (yellow), 6 kpc (orange), 8 kpc (red), 10 kpc (violet), 12 kpc (indigo), and 14 kpc (blue).}
    \label{fig:sfhs}
    \script{star_formation_histories.py}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.32\linewidth]{figures/onezone_powerlaw_slope.pdf}
    \includegraphics[width=0.32\linewidth]{figures/onezone_exponential_timescale.pdf}
    \includegraphics[width=0.32\linewidth]{figures/onezone_plateau_width.pdf}
    \caption{\textit{Left:} Abundance tracks in the [Fe/H]--[O/Fe] plane for one-zone chemical evolution models models which assume a power-law DTD. The dotted, dashed, and solid curves represent models with power-law slopes $\alpha=-0.8$, $-1.1$, and $-1.4$, respectively. In all cases a minimum SN Ia delay time of 40 Myr is assumed. The open symbols located along each curve mark logarithmic steps in simulation time. The marginal panels above and to the right of the central panel present the distributions of [Fe/H] and [O/Fe], respectively. For display purposes these distributions are convolved with a Gaussian with a standard deviation of 0.02 dex.
    \textit{Middle:} A similar plot comparing exponential DTDs with varying timescales. The solid, dashed, and dotted curves represent timescales $\tau=6$, 3, and 1.5 Gyr, respectively.
    \textit{Right:} Similar, comparing plateau DTDs with varying plateau widths. The solid, dashed, and dotted green curves represent widths $W=1$, 0.3, and 0.1 Gyr, respectively. All plateau DTD models assume a post-plateau power-law slope of $\alpha=-1.1$. For reference, the dotted blue curve represents an exponential DTD with $\tau=3$ Gyr, and the dotted purple curve represents a power-law DTD $\alpha=-1.1$ and no plateau.}
    \label{fig:onezone-threepanel}
    \script{onezone_threepanel.py}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.49\linewidth]{figures/onezone_minimum_delay.pdf}
    \includegraphics[width=0.49\linewidth]{figures/onezone_dtd.pdf}
    \caption{\textit{Left:} Comparison of one-zone models with minimum SN Ia delay times of 40 Myr (solid) and 150 Myr (dashed curves) for two DTDs: a power-law with $\alpha=-1.1$ (blue), and an exponential with $\tau=1.5$ Gyr (purple curves). The layout is similar to the plots in Figure \ref{fig:onezone-threepanel}.
    \textit{Right:} Comparison of one-zone models which assume all five functional forms for the DTD in this paper: triple-system (green), plateau with $W=1$ Gyr (cyan), exponential with $\tau=1.5$ Gyr (purple), power-law with $\alpha=-1.1$ (blue), and two-population (pink). In all cases a minimum SN Ia delay time of 40 Myr is assumed.}
    \label{fig:onezone-twopanel}
    \script{onezone_twopanel.py}
\end{figure*}

@@

Holding the SFH fixed, varying the DTD has only a minor effect on the MDF across the disk. Figure \ref{fig:feh-df-dtd} plots the [Fe/H] distributions for two multi-zone simulations which both assume an inside-out SFH but different DTDs: a power-law with slope $\alpha=-1.4$, and an exponential with timescale $\tau=3$ Gyr. The balance between prompt and delayed SNe Ia is starkly different between the two models, with $\sim 80\%$ of SNe Ia exploding within 1 Gyr in the former model but only $\sim 30\%$ in the latter. Nevertheless, the effect of the DTD on the MDF is second-order compared to the effect of the SFH as seen in Figure \ref{fig:feh-df-sfh}.

@@

Figure \ref{fig:feh-df-sfh} plots the distribution of [Fe/H] in multiple regions across the disk for the four SFHs in this paper.

@@

@@

@@

@@

@@

@@

\citet{Greggio2005-AnalyticalRates} derives analytical DTDs for SD and DD progenitor systems from assumptions about binary stellar evolution and outcomes of mass exchange. They find that the parameters which have a large effect on the shape of the DTD are the distribution and range of stellar masses in progenitor systems; the efficiency of accretion in the SD scenario; and the distribution of separations at birth in the DD scenario. Figure \ref{fig:dtd-greggio05} shows the analytical DTDs for SD progenitors and two different prescriptions for DD progenitors (``WIDE'' and ``CLOSE''). In the ``WIDE'' scheme, it is assumed that there is a wide distribution of ratios $A/A_0$ of the separation of the DD system to the initial separation of the binary, and that the distributions of $A$ and total mass of the system $m_{\rm DD}$ are independent, so one cannot necessarily predict the total merge time of a system based on its initial parameters. In the ``CLOSE'' scheme, there is assumed to be a narrow distribution of $A/A_0$ and a correlation between $A$ and $m_{\rm DD}$, so the most massive binaries tend to merge quickly and the least massive merge last.

Figure \ref{fig:onezone-greggio05} shows the results of one-zone chemical evolution models with the \citet{Greggio2005-AnalyticalRates} DTDs. We assume $\eta=2.5$, $\tau_*=2$ Gyr, an inside-out SFH evaluated at a radius of 8 kpc, a continuous recycling approximation, and a minimum SN Ia delay of 40 Myr. We compare the SD, DD WIDE, and DD CLOSE schemes to standard power-law, broken power-law, and exponential DTDs. The SD and DD CLOSE DTDs follow nearly identical tracks in [O/Fe] vs [Fe/H]; however, their distributions on [O/Fe] differ at the low end. The SD DTD follows an exponential with a 1.5 Gyr timescale, whereas the DD CLOSE DTD is well-approximated by a broken power-law with an initial plateau of 300 Myr and a subsequent declining slope of -1.1. The WIDE prescription is likewise best approximated by a broken power-law, but with a longer plateau width of 1 Gyr. In all cases, the difference between the analytical DTD and the nearest broken power-law or exponential is likely too small to be observationally detectable, so in our multi-zone models we implement the more generic functions in lieu of the analytical DTDs.

>>>>

\usepackage{showyourwork}
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{layouts}
\usepackage{xcolor}
\usepackage{upgreek}

@@

Summary of DTDs used in simulations and theoretical studies. The Feedback In Realistic Environments \citep[FIRE;][]{Hopkins2014-FIRE-1} and FIRE-2 \citep{Hopkins2018-FIRE-2} simulations implement the DTD from \citet{Mannucci2006-TwoPopulations} (explain previously). Following \citet{Schonrich2009-RadialMixing}, \citet{Weinberg2017-ChemicalEquilibrium} assume an exponential DTD for the purposes of producing an analytical solution to the equations of chemical equilibrium, but they show that it can be extended to a double-exponential DTD which closely approximates a power-law.

A few chemical evolution studies have compared DTDs with different functional forms. \citet{Poulhazan2018-PrecisionPollution} investigate six different DTDs in a smoothed-particle hydrodynamics simulation. \citet{Palicio2023-AnalyticDTD} derive analytic solutions to one-zone chemical evolution equations for seven DTDs (three theoretical and four empirical), approximating the more complicated DTD functions as a linear combination of exponential, Gaussian, and power-law components.

The origin of the [$\alpha$/Fe] bimodality is debated. \citet{Schonrich2009-RadialMixing} argue that it arises due to radial migration (churning) even with a smooth star formation rate. \citet{Johnson2021-Migration} use a smooth SFR and a prescription for radial migration but are unable to reproduce the bimodality. \citet{Spitoni2021-TwoInfall} (definitely earlier papers too) argue that two distinct infall episodes can explain the bimodality.

@@

\begin{figure}
    \centering
    \includegraphics{figures/delay_time_distributions.pdf}
    \caption{Models of the SN Ia delay time distribution (DTD) used in this paper. All functions are normalized to $f_{\rm Ia}=1$ at $\tau=1$ Gyr. The black squares represent the DTD recovered for the SDSS-II sample of SNe Ia by \citet{Maoz2012-SloanIIDTD}, to the same scale as the DTD models. The horizontal error bars indicate the time range of each DTD measurement, and the vertical error bars represent the 1$\sigma$ uncertainties.}
    \label{fig:dtds}
    \script{delay_time_distributions.py}
\end{figure}

@@

\paragraph{Plateau} A ``plateau'' model consisting of a flat slope of width $W$ followed by a power law decline with normalization $B_W$:
\begin{equation}
    R(t) = 
    \begin{cases}
        B_W, & t < W \\
        B_W (t/W)^\alpha, & t \ge W
    \end{cases}
    \label{eq:plateau-dtd}
\end{equation}
where $B_W=[(W-t_{\rm min}) + \frac{t_{\rm max}^{\alpha+1} - W^{\alpha+1}}{(\alpha+1)W^\alpha}]^{-1}$ is the normalization coefficient.
We investigate the cases $W=0.3$ Gyr and $W=1$ Gyr, taking $\alpha=-1.1$ for all plateau models. Figure \ref{fig:analytical-dtd} illustrates the similarity between these models and two different treatments of the analytical double-degenerate DTD from \citet{Greggio2005-AnalyticalRates}. For more discussion on this, see Appendix \ref{app:analytical-dtds}.

@@

\paragraph{Two-population} A DTD in which 50\% of SNe Ia belong to a ``prompt'' Gaussian component at small $t$ and the other 50\% form an exponential tail at large $t$, normalized by $D_{t_p,\sigma,\tau}$:
\begin{equation}
    R(t) = D_{t_p,\sigma,\tau} \Big[e^{-\frac{(t-t_p)^2}{2\sigma^2}} + e^{-t/\tau}\Big]
    \label{eq:prompt-dtd}
\end{equation}
Following the two-population model from \citet{Mannucci2006-TwoPopulations}, we take $t_p=50$ Myr, $\sigma=15$ Myr, and $\tau=3$ Gyr, which results in roughly 50\% of SNe Ia exploding within $t<100$ Myr. \citet{Poulhazan2018-PrecisionPollution} refer to this as the ``bimodal'' model.

@@

\begin{figure*}
    \centering
    \includegraphics{figures/star_formation_histories.pdf}
    \caption{The surface densities of star formation $\dot \Sigma_*$ (first row), gas infall $\dot \Sigma_{\rm in}$ (second row), and gas mass $\Sigma_{\rm gas}$ (third row), and the star formation efficiency timescale $\tau_*$ (fourth row) as functions of simulation time for our four model SFHs: inside-out (first column; see Equation \ref{eq:insideout-sfh}), late-burst (second column; see Equation \ref{eq:lateburst-sfh}), early-burst (third column; see Equations \ref{eq:earlyburst-ifr} and \ref{eq:earlyburst-taustar}), and two-infall (fourth column; see Equation \ref{eq:twoinfall-ifr}). In each panel, we plot curves for the simulation zones which have inner radii at 4 kpc (yellow), 6 kpc (orange), 8 kpc (red), 10 kpc (violet), 12 kpc (indigo), and 14 kpc (blue).}
    \label{fig:sfhs}
    \script{star_formation_histories.py}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.32\linewidth]{figures/onezone_powerlaw_slope.pdf}
    \includegraphics[width=0.32\linewidth]{figures/onezone_exponential_timescale.pdf}
    \includegraphics[width=0.32\linewidth]{figures/onezone_plateau_width.pdf}
    \caption{\textit{Left:} Abundance tracks in the [Fe/H]--[O/Fe] plane for one-zone chemical evolution models models which assume a power-law DTD. The dotted, dashed, and solid curves represent models with power-law slopes $\alpha=-0.8$, $-1.1$, and $-1.4$, respectively. For reference, the solid gray curve represents an exponential DTD with $\tau=3$ Gyr. In all cases a minimum SN Ia delay time of 40 Myr is assumed. The open symbols located along each curve mark logarithmic steps in simulation time. The marginal panels above and to the right of the central panel present the distributions of [Fe/H] and [O/Fe], respectively. For display purposes, these distributions are convolved with a Gaussian with a standard deviation of 0.02 dex.
    \textit{Middle:} Similar, comparing exponential DTDs with varying timescales. The solid, dashed, and dotted curves represent timescales $\tau=6$, 3, and 1.5 Gyr, respectively.
    \textit{Right:} Similar, comparing plateau DTDs with varying plateau widths. The solid, dashed, and dotted green curves represent widths $W=1$, 0.3, and 0.1 Gyr, respectively. All plateau DTD models assume a post-plateau power-law slope of $\alpha=-1.1$. For reference, the solid gray curve represents an exponential DTD with $\tau=3$ Gyr, and the dotted purple curve represents a power-law DTD $\alpha=-1.1$ and no plateau.}
    \label{fig:onezone-threepanel}
    \script{onezone_threepanel.py}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.49\linewidth]{figures/onezone_minimum_delay.pdf}
    \includegraphics[width=0.49\linewidth]{figures/onezone_dtd.pdf}
    \caption{\textit{Left:} Comparison of one-zone models with minimum SN Ia delay times of 40 Myr (solid curves) and 150 Myr (dashed curves) for two DTDs: a power-law with $\alpha=-1.1$ (blue curves), and an exponential with $\tau=1.5$ Gyr (purple curves). The layout is similar to the plots in Figure \ref{fig:onezone-threepanel}.
    \textit{Right:} Comparison of one-zone models which assume all five functional forms for the DTD in this paper: triple-system (green), plateau with $W=1$ Gyr (cyan), exponential with $\tau=1.5$ Gyr (purple), power-law with $\alpha=-1.1$ (blue), and two-population (pink). A minimum SN Ia delay time of 40 Myr is assumed for all models.}
    \label{fig:onezone-twopanel}
    \script{onezone_twopanel.py}
\end{figure*}

@@

Figure \ref{fig:feh-df-comparison} shows MDFs across the Galaxy for a selection of models and APOGEE data. The two left-hand columns show MDFs from our simulations which assume the same DTD (exponential with $\tau=1.5$ Gyr) but different model SFHs. For comparison, the APOGEE MDFs are binned similarly and presented in the center column of Figure \ref{fig:feh-df-comparison}.

Holding the SFH fixed, varying the DTD has a very minor effect on the MDF across the disk. The two right-hand columns of Figure \ref{fig:feh-df-comparison} plot the MDFs for two multi-zone simulations which both assume an inside-out SFH but different DTDs: a power-law with slope $\alpha=-1.4$, and an exponential with timescale $\tau=3$ Gyr. The balance between prompt and delayed SNe Ia is starkly different between the two models, with $\sim 80\%$ of SNe Ia exploding within 1 Gyr in the former model but only $\sim 30\%$ in the latter.

The choice of DTD and SFH has a larger effect on the inner regions of the galaxy than the outer. Inner regions have a higher SFR and more stars are formed at early times than late, so differences in the SN Ia rate have a more pronounced effect on the MDF. For all SFH models except two-infall, the outer galaxy has a longer SFR timescale than the inner galaxy, so the MDF in the outer galaxy is relatively unaffected by the fraction of prompt SNe Ia.

While there are some quantitative differences in how the shape of the MDF varies with Galactic region, the qualitative trends are unaffected by the choice of model SFH or DTD. These trends are primarily driven by chemical equilibrium, abundance gradients, and radial migration \citep[][see their Section 3.2 for further discussion]{Johnson2021-Migration}.

@@

@@

The early-burst model produces a bimodal distribution of [O/Fe] even out to large radii and close to the midplane, whereas the APOGEE distribution does not show a prominent high-$\alpha$ peak beyond $R_{\rm{Gal}}\sim11$ kpc.

@@

Figure \ref{fig:ofe-df-dtd} shows [O/Fe] distributions from simulations with the same SFH but a variety of different DTDs. We choose the early-burst SFH because it produces distinct high- and low-$\alpha$ sequences, which makes the effect of the DTD on the former more apparent. The most apparent effect of the DTD is to shift the mode of the high-$\alpha$ sequence. The two-population DTD, which has the most prompt SNe Ia of any model in this paper, places the high-$\alpha$ sequence at [O/Fe]$\approx 0.2$, while the triple-system DTD, which has the fewest prompt SNe Ia, places $\sim0.2$ dex higher at [O/Fe]$\approx0.4$.

@@

\subsection{Bimodality in [O/Fe]}
\label{sec:bimodality}

@@

Figure \ref{fig:ofe-feh-sfh} compares the [O/Fe]--[Fe/H] plane in the Solar neighborhood ($7\leq R_{\rm gal}<9$ kpc, $0\leq|z|<0.5$ kpc) between our four model SFHs.

@@

Compare to previous studies: \citet{Poulhazan2018-PrecisionPollution,Palicio2023-AnalyticDTD}.

\citet{Poulhazan2018-PrecisionPollution} found that DTD models with a significant prompt component (power-law and two-population) produced the highest average stellar abundance of [O/Fe]. They also found that these models produced the narrowest [O/Fe] distributions and suppressed the long tail to low [O/Fe] ratios. Conversely, their models with essentially 0 prompt SNe Ia produced a lower average stellar [O/Fe] ratio at late times and a broader distribution of [O/Fe]. Their isolated galaxy simulations have an SFR which features an early ($<1$ Gyr) burst followed by a roughly exponential decline, qualitatively similar to our early-burst model but with the peak of star formation occuring $\sim3$ Gyr earlier. We note that their simulations were not designed to reproduce a Milky Way-like galaxy; as such, their [O/Fe] distributions peak around $\sim0.45$ which is higher than the APOGEE high-$\alpha$ sequence. Nevertheless, our study reproduces their finding that a DTD with more prompt SNe Ia will produce a narrower distribution of [O/Fe]. One exception to this is the exponential DTD, which has a marginally narrower distribution than the power-law DTD close to the midplane [numbers here] but a more prominent hith-$\alpha$ tail. This is likely a consequence of the fact that the exponential DTD has a much lower SN Ia rate at late times ($\sim10$ Gyr) than other models [expand on this].

@@

\citet{Greggio2005-AnalyticalRates} derives analytical DTDs for SD and DD progenitor systems from assumptions about binary stellar evolution and outcomes of mass exchange. They find that the parameters which have a large effect on the shape of the DTD are the distribution and range of stellar masses in progenitor systems; the efficiency of accretion in the SD scenario; and the distribution of separations at birth in the DD scenario. The left-hand panel of Figure \ref{fig:analytical-dtd} shows the analytical DTDs for SD progenitors and two different prescriptions for DD progenitors (``WIDE'' and ``CLOSE''). In the ``WIDE'' scheme, it is assumed that there is a wide distribution of ratios $A/A_0$ of the separation of the DD system to the initial separation of the binary, and that the distributions of $A$ and total mass of the system $m_{\rm DD}$ are independent, so one cannot necessarily predict the total merge time of a system based on its initial parameters. In the ``CLOSE'' scheme, there is assumed to be a narrow distribution of $A/A_0$ and a correlation between $A$ and $m_{\rm DD}$, so the most massive binaries tend to merge quickly and the least massive merge last.

The right-hand panel of Figure \ref{fig:analytical-dtd} shows the results of one-zone chemical evolution models with the \citet{Greggio2005-AnalyticalRates} DTDs. We assume $\eta=2.5$, $\tau_*=2$ Gyr, an inside-out SFH evaluated at a radius of 8 kpc, a continuous recycling approximation, and a minimum SN Ia delay of 40 Myr. We compare the SD, DD WIDE, and DD CLOSE schemes to standard power-law, broken power-law, and exponential DTDs. The SD and DD CLOSE DTDs follow nearly identical tracks in [O/Fe] vs [Fe/H]; however, their distributions on [O/Fe] differ at the low end. The SD DTD follows an exponential with a 1.5 Gyr timescale, whereas the DD CLOSE DTD is well-approximated by a broken power-law with an initial plateau of 300 Myr and a subsequent declining slope of -1.1. The WIDE prescription is likewise best approximated by a broken power-law, but with a longer plateau width of 1 Gyr. In all cases, the difference between the analytical DTD and the nearest broken power-law or exponential is likely too small to be observationally detectable, so in our multi-zone models we implement the more generic functions in lieu of the analytical DTDs.
